---
title: "Template"
author: "Studentnames and studentnumbers here"
date: "`r Sys.Date()`"
output:
  pdf_document: default
---

# Set-up your environment

```{r package_install, include=FALSE}
if (!requireNamespace("tidyverse", quietly = TRUE)) install.packages("tidyverse")
if (!requireNamespace("yaml", quietly = TRUE)) install.packages("yaml")
if (!requireNamespace("rmarkdown", quietly = TRUE)) install.packages("rmarkdown")
if (!requireNamespace("dplyr", quietly = TRUE)) install.packages("dplyr")
if (!requireNamespace("cbsodataR", quietly = TRUE)) install.packages("cbsodataR")
if (!requireNamespace("sf", quietly = TRUE)) install.packages("sf")
if (!requireNamespace("readr", quietly = TRUE)) install.packages("readr")
if (!requireNamespace("ggplot2", quietly = TRUE)) install.packages("ggplot2")
```

```{r packages}
library(tidyverse)
library(rmarkdown)
library(yaml)
library(dplyr)
library(cbsodataR)
library(sf)
library(readr)
library(ggplot2)
```

# Title Page

Include your names

Include the tutorial group number

Include your tutorial lecturer's name

# Part 1 - Identify a Social Problem

Use APA referencing throughout your document. [Here's a link to some explanation.](https://www.mendeley.com/guides/apa-citation-guide/)

## 1.1 Describe the Social Problem

Include the following:

-   Why is this relevant?

-   ...

# Part 2 - Data Sourcing

## 2.1 Load in the data

Preferably from a URL, but if not, make sure to download the data and store it in a shared location that you can load the data in from. Do not store the data in a folder you include in the Github repository!

```{r loading_data}
Welzijn_Goed <- read_csv("Welzijn.goed.csv")

Ervaren_Gezondheid <- read_csv("Ervaren_Gezondheid.csv")

Levensverwachting <- read_csv("Levensverwachting.csv")

Ervaren_gezondheid_wijk <- read_delim("Ervarengezondheid_Wijk&Buurt.csv", delim = ";")

Inkomen_per_gemeente <- read_delim("Inkomen_gemeente.csv", delim = ";")

Levensverwachting_Gemeente <- read_delim("Levensverwacht_Gemeente_Wijk&Buurt.csv", delim = ";")

ErvarenGezondheidNL <- read_csv("ErvarenGezondheidNL.csv")

gemeentegrenzen <- st_read("https://service.pdok.nl/cbs/gebiedsindelingen/2023/wfs/v1_0?request=GetFeature&service=WFS&version=2.0.0&typeName=gemeente_gegeneraliseerd&outputFormat=json")
```

midwest is an example dataset included in the tidyverse package

## 2.2 Provide a short summary of the dataset(s)

```{r}
head(dataset)
```

In this case we see 28 variables, but we miss some information on what units they are in. We also don't know anything about the year/moment in which this data has been captured.

URL van de data download pagina

``` r
inline_code = TRUE
```

These are things that are usually included in the metadata of the dataset. For your project, you need to provide us with the information from your metadata that we need to understand your dataset of choice.

## 2.3 Describe the type of variables included

Think of things like:

-   Do the variables contain health information or SES information?

-   Have they been measured by interviewing individuals or is the data coming from administrative sources?

*For the sake of this example, I will continue with the assignment...*

# Part 3 - Quantifying

## 3.1 Data cleaning

Say we want to include only larger distances (above 2) in our dataset, we can filter for this.

```{r data_cleaning}
mean(dataset$percollege)
```

Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.

## 3.2 Generate necessary variables

Variable 1

```{r gen_var1}
gemeente_cor <- gemeente_wijk_data %>%
  group_by(statnaam) %>%
  filter(n() >= 3) %>%
  summarise(
    correlation = if (sum(complete.cases(as.numeric(Gemiddeld, ErvarenGezondheidGoedZeerGoed_4))) > 1)
      {cor(as.numeric(Gemiddeld), as.numeric(ErvarenGezondheidGoedZeerGoed_4), use = "complete.obs")} 
    else {NA_real_})

###!!! CHECK IF THIS PART BELOW HAS TO BE EVEN IN HERE!!!###
#This isnt creating new var, but adding to new data set
#trimming white spaces so I can add the geometry for the heat map
gemeentegrenzen <- gemeentegrenzen %>%
  mutate(statnaam = tolower(trimws(statnaam)))
gemeente_cor <- gemeente_cor %>%
  mutate(statnaam = tolower(trimws(statnaam)))

gemeente_wijk_mapdata <- gemeentegrenzen %>%
  left_join(gemeente_cor, by = "statnaam")
```

We decided to make a new variable called "correlation" that we added

Variable 2

```{r gen_var2}

```

## 3.3 Visualize temporal variation

```{r}
ggplot(Ervaren_Gezondheid_NL_AM_RO, 
       aes(x = Perioden, y = ErvarenGezondheidGoedZeerGoed_4,
           colour = Region, group = Region)) +
  geom_line(size = 1.5) +
  geom_point(colour = "black", size = 2.5, show.legend = FALSE) +
  geom_text(aes(label = ErvarenGezondheidGoedZeerGoed_4,
                vjust = ifelse(Region == "The Netherlands", -0.8, 1.9)),
            colour = "black", 
            show.legend = FALSE) +
  scale_colour_manual(values = c(
    "Amsterdam" = "#C41230",
    "Rotterdam" = "#39B54A",
    "The Netherlands" = "orange"
  )) +
  scale_x_discrete(labels = c("2012", "2016", "2020")) +
  scale_y_continuous(limits = c(65, 80), breaks = seq(0, 100, by = 5)) +
  labs(
    x = "Year (2012, 2016, 2020)",
    y = "Perceived health (% wel or very well)",
    title = "Perceived Health per Municipality",
    colour = "Region"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## 3.4 Visualize spatial variation

```{r visualise_map}
ggplot(gemeente_wijk_mapdata) +
  geom_sf(aes(fill = correlation)) + 
  theme_minimal() +
  scale_fill_gradient2(high = "#6A0DAD", mid = "lightgrey", low = "#FFD700",
                       limits = c(-1, 1),
                       breaks = c(-1, -0.5, 0, 0.5, 1),
                       name = "Correlation 'income & health'") +
  theme(legend.position = "left") + 
  labs(caption = "Grey areas = insufficient data") 

```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations}
ggplot(Welzijn_temporal_visualization, aes(x = Jaar, y = WelzijnIndex, color = Opleidingsniveau, group = Opleidingsniveau)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = 2015:2018) +
  scale_color_manual(
    values = c(
      "Basisonderwijs" = "#F8766D",
      "Vmbo, havo-, vwo-onderbouw, mbo1" = "#C49A00",
      "Havo, vwo, mbo2-4" = "#00BA38",
      "Hbo-, wo-bachelor" = "#00BFC4",
      "Hbo-, wo-master, doctor" = "#C77CFF"
    )
  ) +
  labs(
    title = "Evolution of Welfare Index per Level of Education (2015–2018)",
    x = "Year",
    y = "Welfare Index",
    color = "Education level"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 9, face = "bold", hjust = 0.5),
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10))
  )
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

https://github.com/TijsReijling/Programmeren 

## 5.2 Reference list

#Bovenaan welvaartsladder bijna 25 jaar langer in goede gezondheid. (2022, December 20). Centraal Bureau Voor De Statistiek. <https://www.cbs.nl/nl-nl/nieuws/2022/51/bovenaan-welvaartsladder-bijna-25-jaar-langer-in-goede-gezondheid>?

#Pharos. (2022, July). Sociaal economische Gezondheidsverschillen (SEGV). <https://www.pharos.nl/factsheets/sociaaleconomische-gezondheidsverschillen-segv/>

#Sociaaleconomische gezondheidsverschillen \| Volksgezondheid en Zorg. (n.d.). <https://www.vzinfo.nl/sociaaleconomische-gezondheidsverschillen>

These sources include links to the datasets and their corresponding metadata.

Gezonde levensverwachting; inkomen en welvaart. (2025, May 23). CBS dataportaal. <https://opendata.cbs.nl/statline/portal.html?_la=nl&_catalog=CBS&tableId=85445NED&_theme=154>

Welzijn; kerncijfers, persoonskenmerken. (2025, March 20). CBS dataportaal. <https://opendata.cbs.nl/statline/portal.html?_la=nl&_catalog=CBS&tableId=85542NED&_theme=178>

Inkomen per gemeente en wijk, 2020. (2023, September 1). Centraal Bureau voor de Statisktiek. <https://www.cbs.nl/nl-nl/maatwerk/2023/35/inkomen-per-gemeente-en-wijk-2020> (metadata included in dataset)

Gezondheid per wijk en buurt; 2012/2016/2020/2022. (2024, December 09). RIVM dataportaal. <https://statline.rivm.nl/portal.html?_la=nl&_catalog=RIVM&tableId=50120NED&_theme=94>

Levensverwachting op de leeftijd 0 en 65 jaar; geslacht, regio 1996-2022. (2024, December 9). RIVM dataportaal. <https://statline.rivm.nl/portal.html?_la=nl&_catalog=RIVM&tableId=50132NED&_theme=101>
